### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/response/response.tex

\itemitem{} Thomas C.H. Lux, Layne T. Watson, et al: An Algorithm for Constructing Monotone Quintic Interpolating Splines. Proceedings of the 2020 Spring Simulation Conference, May 2020, Article No. 33, Pages 1-12.

\item{} The differences with respect to the current manuscript seem to be rather specific and minor, and at least I would have expected a discussion/comparison between the current approach and the one in the above published work.

}

That conference paper was an earlier version of this work and may appear similar at first glance, but there are many critical differences between the two (computations refactored for numerical stability, different method for estimating derivatives, using a binary search to find feasible derivative values for the entire spline rather than a single interval). Notably, there was no code published with that conference paper, which is one of the most important components of this submission. To make these differences more clear, two sentences were added to the end of paragraph 4 in section 1.

@@

\item{(3)} The literature study on monotone spline interpolation is incomplete and misses important work. As far as I am aware of, one of the most commonly used methods for shape-preserving interpolation is the quadratic spline method by Larry Schumaker (it does not only preserves monotonicity but also convexity!), at least in the spline approximation theory community:

\itemitem{} Larry L. Schumaker: On Shape Preserving Quadratic Spline Interpolation. SIAM Journal on Numerical Analysis, Vol. 20, No. 4 (1983), Pages 854-864.

}

The submission cites quadratic, cubic, and quartic previous works. The work by Larry Schumaker can readily be included in this list, and has been included in the attached update to the paper (section 1, paragraph 3). It is our understanding that one of the most widely used shape-preserving spline algorithms is a cubic spline methodology proposed by Fritsch and Carlson (known as PCHIP), which is included in many mathematical software libraries and cited in the submission.

@@

{\parindent=20pt \it

\item{(4)} The numerical section just shows few examples of data sets and the obtained spline results. The authors make some general statements about the algorithm like ``it generally provides aesthetically pleasing (low wiggle) initial estimates ... is preferred because it results in monotone quintic spline interpolants that are nearer to initial estimates'' but a deeper (more objective) discussion is lacking. Also a numerical comparison with other classical monotone spline methods is missing, at least the algorithm by Schumaker [1983] and also the one by Fritz/Carlson [1980]. This might help understand why the new method is of interest.

}

This is a reasonable criticism, but also easily addressed with minor revisions to the submission. While the reviewer may disagree with the language we used, there are objective reasons for the decisions made in this work in addition to the subjective comments made. Specifically, the quadratic facet model picks the minimum curvature local estimate for second derivatives which has the smallest magnitude second derivative, and the binary search identifies derivative values absolutely nearer to the initial estimates than the previous method of shrinking derivative values used in (1). These facts are now mentioned in the paper (section 4, paragraphs 2 and 3).

Comparisons with lower order methods are not appropriate for this work. There is no published mathematical software that also constructs a monotone quintic spline interpolant to data. If a user desires a monotonicity preserving $C^2$ approximation, this would be the only published software that achieves that goal. This paper is not intended to be a review and comparison of all available shape preserving spline packages. Including such comparisons would also require subjective decisions about what visuals to show, and those decisions might draw criticisms that distract from the unique purpose and achievement of this work (producing $C^2$ monotonicity preserving approximations to data). To emphasize that this work is not a replacement for previous works, but rather an additional method to consider, the first sentence of section 1, paragraph 5 that mentioned the importance of this work ``over'' previous methods was removed.

@@

}

The reviewer may have missed the reason for the global linear system, from the submission (section 5, paragraph 4), ``It would be faster to construct splines over intervals independently ... the single linear system is chosen here for the decreased redundancy''. To be clear, the de Boor-Cox algorithm {\it is} used to evaluate the underlying B-splines (section 5, paragraph 1). The linear combination in the submission is of the (quintic) B-splines that define the interpolant.

\bigskip

To the best of our knowledge, rejection based on a single review has not been a journal policy and we respectfully request additional reviews. We believe this algorithm and the accompanying code has great potential for TOMS, noting that the quadratic and cubic works preceding this have remained state-of-the-art for forty years. Thank you for both your time and consideration.

### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Next round of response done, sending to coauthors.

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations with
large {\it wiggle}, or whose second derivative functions have large
$L2$ norm over the domain. In an attempt to capture the local shape of
the data while minimizing wiggle, this package uses a facet model from
image processing [Haralick and Watson 1981] to estimate first and
second derivatives at breakpoints. Rather than picking a local linear
or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.

>>>>

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large $L2$ norm (approximations
with large {\it wiggle}). In an attempt to capture the local shape of
the data while minimizing wiggle, this package uses a facet model from
image processing [Haralick and Watson 1981] to estimate first and
second derivatives at breakpoints. Rather than picking a local linear
or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.
